An Accurate Model for SARS-CoV-2 Pooled RT-PCR Test Errors

### 2021-09-12 paper.tex
intro and minor changes everywhere

\section*{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in breaking transmission chains and mitigating the
COVID-19 pandemic. As such, the need for large-scale testing has
resulted in the application of techniques to increase the throughput
of RT-PCR tests \cite{DorfmanYuvalDor, PoolSize30, BayesianDorfman,
  MatrixPooling, LionDorfman, CherifReview}. These techniques, often
termed \emph{pooling}, or group testing, have originated in the 1940's
in the seminal work of Dorfman and his pooling technique
\cite{DorfmanOriginal, DorfmanYuvalDor}. Under Dorfman pooling, one
selects $N$ individuals and performs a single RT-PCR test on their
combined (\emph{pooled}) samples. If the pooled test yields a positive
result, then each individual is retested separately; otherwise,
everyone is declared negative. The throughput efficiency of Dorfman
pooling has been demonstrated empirically \cite{DorfmanYuvalDor} and
its error rates thoroughly investigated \cite{Kim, Simplistic1,
  OptimalDorfmanPool}.

The study and practice of pooling has advanced considerably since Dorfman's original paper, with 
several other pooling techniques being applied for SARS-CoV-2.
For example, in \emph{recursive} pooling \cite{Kim,RecursiveSevenFold}, if the
first pooled test is positive, the pool is split into two and the
process repeats. Otherwise, all pool members are declared negative.
In \emph{Matrix} pooling \cite{MatrixPooling} a population of size
$N=mn$ is arranged in a matrix of dimensions $m\times n$. Each
row and column are then pooled and individuals in the intersection of
positive rows and columns are tested separately. Other, more sophisticated
pooling schemes exist \cite{CompressedPooling}, but studies investigating the 
underlying assumptions of pooling are surprisingly lacking.

Studies focused on pooling for SARS-CoV-2 are in a consensus that
sample dilution effects \cite{DilutionHIV, GroupDilution} are not a
concern, even for pools as large as 64 individuals \cite{PoolSize30,
  Lion, DorfmanYuvalDor, DilutionCOVID, CherifReview}. Consequently,
studies assumed that the probability of a true-positive (the test's
\emph{sensitivity}) does not depend on the number of infected samples
in the pool, but rather on the existence of at least one such
sample. Thus, the probability of a positive result in a pooled test
has been (assumed) identical for a pool with one sample from an
infected individual and, e.g., five such samples. This assumption is
common in the group testing literature \cite{Kim, OptimalDorfmanPool,
  CherifReview}, as well as in more specific, COVID-19 focused studies
\cite{Simplistic1, Simplistic2}. In this study we challenge this
commonly made assumption and show how using a more accurate
probabilistic model affects the estimation of false-negative rates for
Dorfman pooling.

Several consequences arise from our analysis: Previously estimated optimal Dorfman
pool size should be revisited, as it is influenced by the prevalence \cite{OptimalDorfmanPool}. The error
rates of many pooling schemes in use should be reesetimated
\cite{BayesianDorfman, Kim}. And, finally, new methods should be
developed in order to take advantage of our novel model \cite{DOPE}.

@@

\section*{Results}\label{section:results}
\subsection*{Scheme false-negative}
We plot $\err{exact}$ for varying prevalence $q$ and sensitivity $\Se$
values, and make the comparison with $\err{common}$. As recommended by
\cite{DorfmanYuvalDor}, we apply different pool sizes $N$, for
different prevalence values. We observe that for a false-positive rate
$\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of reasonable
sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}, the increase in scheme false-negative rate
(compared to $1-\Se$), summarized in $\err{exact}$, is at leaset
$60\%$ (Figure \ref{fig1}).

Interestingly, an increase in infection prevalence monotonically
decreases the scheme false-negative rate, as can also be easily seen
from equation \eqref{eq:sfn}. For the chosen parameter ranges, the increase in
the single test false-negative rates increases the relative error
$\err{exact}$. These effects can be seen in Figure~\ref{fig1} (left
panel), upon conditioning on pool size. Extending the range for $\Sp$
yields no qualitative differences. We further compare $\err{exact}$ to
$\err{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure~\ref{fig1}, right
panel).
\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{heatmap_sfn.jpg}
  \caption{Relative increase in Dorfman pooling false-negative rates
    $\err{exact}$. Left: Colors represent $\err{exact}$, the relative
    percentage increase in the scheme false-negative rates relative to
    the single test false-negative rates equation \eqref{eq:erel}. Right:
    colors represent the difference between $\err{common}$ and
    $\err{exact}$. The disease prevalence $q$, is varied on the
    x-axis, while the test sensitivity is varied on the y-axis. Pool
    size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}. The left panel shows that $\err{exact}$ is
    largest for low prevalence values $q$. The difference between
    $\err{common}$ and $\err{exact}$ can be as large as 30\%, as seen
    in the right panel. }\label{fig1}
\end{figure}

@@

Furthermore, an especially important consideration in designing
pooling schemes is explicitly taking the intrinsic RT-PCR error rates
into account. As we have demonstrated above, analyzing the expected
error rates of pooling schemes under a realistic error
model is readily achievable. Designing pooling schemes which optimize
testing while accounting for error rates is, however, considerably more involved.
Nonetheless, various algorithms explicitly incorporate error
rates into their pooling optimization processes \cite{Kim, OptimalDorfmanPool}, including a
recent study by our group which employed the herein described error
model \cite{DOPE}.  In our proposed algorithm, we have used an
experimental design criterion to create an algorithm which iteratively
performs pooling, updating its information based on previous
results. In doing so, our pooling scheme shows potential in terms of
both reducing false-negative and false-positive rates, as well as increasing efficiency,
although it has not yet been implemented in real-world settings.

>>>>

\section*{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in the fight against the COVID-19 pandemic. PCR testing can be performed on individuals suspected of infection and hence break potential transmission chains. Moreover, in some countries, PCR testing has become a routine procedure of surveillance, testing all incoming air passengers in airports (ref), schools students (ref), or even the entire adult population (ref).  
As such, the need for large-scale testing has
resulted in the application of techniques to increase the throughput
of RT-PCR tests \cite{DorfmanYuvalDor, PoolSize30, BayesianDorfman,
  MatrixPooling, LionDorfman, CherifReview}. These techniques, often
termed \emph{pooling}, or group testing, have originated in the 1940's
in the seminal work of Dorfman and his pooling technique
\cite{DorfmanOriginal, DorfmanYuvalDor}. Under Dorfman pooling, one
selects $N$ individuals and performs a single RT-PCR test on their
combined (\emph{pooled}) samples. If the pooled test yields a positive
result, then each individual is retested separately; otherwise,
everyone is declared negative. The throughput efficiency of Dorfman
pooling has been demonstrated empirically \cite{DorfmanYuvalDor} and
its error rates thoroughly investigated \cite{Kim, Simplistic1,
  OptimalDorfmanPool}.

The study and practice of pooling has advanced considerably since Dorfman's original paper, with 
several other pooling techniques being applied for SARS-CoV-2.
For example, in \emph{recursive} pooling \cite{Kim,RecursiveSevenFold}, if the
first pooled test is positive, the pool is split into two and the splitting
process repeats. Otherwise, all pool members are declared negative.
In \emph{Matrix} pooling \cite{MatrixPooling} a population of size
$N=mn$ is arranged in a matrix of dimensions $m\times n$. Each
row and column are then pooled and individuals in the intersection of
positive rows and columns are tested separately. Other, more sophisticated
pooling schemes exist \cite{CompressedPooling}, but studies investigating the 
underlying assumptions of pooling are surprisingly lacking.

Studies focused on pooling for SARS-CoV-2 are in a consensus that
sample dilution effects \cite{DilutionHIV, GroupDilution} are not a
concern, even for pools as large as 64 individuals \cite{PoolSize30,
  Lion, DorfmanYuvalDor, DilutionCOVID, CherifReview}. Consequently,
studies assumed that the probability of a true-positive (the test's
\emph{sensitivity}) does not depend on the number of infected samples
in the pool, but rather on the existence of at least one such
sample. Thus, the probability of a positive result in a pooled test
has been (assumed) identical for a pool with one sample from an
infected individual and, e.g., five such samples. This assumption is
common in the group testing literature \cite{Kim, OptimalDorfmanPool,
  CherifReview}, as well as in more specific, SARS-CoV-2 focused studies
\cite{Simplistic1, Simplistic2}. In this study we challenge this
commonly made assumption and demostrate how using a more accurate
probabilistic model affects the estimation of false-negative rates for
Dorfman pooling.

Several consequences arise from our analysis: Previously estimated optimal Dorfman
pool size \cite{OptimalDorfmanPool} should be revisited in the case of SARS-CoV-2, as it is influenced by the prevalence of infection. The error
rates of other pooling schemes in use should also be reesetimated
\cite{BayesianDorfman, Kim}. Another consequence of our results is that studies that seek precise error models should be conducted for other pathogens as well, since the standard error model could prove flawed for other pathogens. Specifically, all assumptions should be scrutinized, including (e.g.) the independence of tests, an  extremely widespread assumption in the group testing literature \cite{Kim, OptimalDorfmanPool} which dates back to Dorfman's original paper \cite{DorfmanOriginal}. Finally, new methods should be developed in order to take advantage of our (and other) more realistic models.

@@

\section*{Results}\label{section:results}
\subsection*{Scheme false-negative}
We plot $\err{exact}$ for varying prevalence $q$ and sensitivity $\Se$
values, and make the comparison with $\err{common}$. As recommended by
\cite{DorfmanYuvalDor}, we apply different pool sizes $N$, for
different prevalence values. We observe that for a false-positive rate
$\Sp=0.95$ \cite{DorfmanYuvalDor} and a range of reasonable
sensitivity and prevalence values \cite{KitComparison,
  InterpretingCOVID19Test, EstimatingRatesLourenco,
  FalsePositiveEstimate}, the increase in the scheme false-negative rate
(compared to $1-\Se$), expressed by $\err{exact}$, is at leaset
$60\%$ (Figure \ref{fig1}).

Interestingly, an increase in infection prevalence monotonically
decreases the scheme false-negative rate, as can also be easily seen
from equation \eqref{eq:sfn}. For the chosen parameter ranges, the increase in
the single test false-negative rates increases the relative error
$\err{exact}$. These effects can be seen in Figure~\ref{fig1} (left
panel), upon conditioning on pool size. Extending the range for $\Sp$
yields no qualitative differences. We further compare $\err{exact}$ to
$\err{common}$, showing the discrepancy changes as a function of both
prevalence and the single test sensitivity (Figure~\ref{fig1}, right
panel).
\begin{figure}[H]
  \centering
  \includegraphics[width=\textwidth]{heatmap_sfn.jpg}
  \caption{Relative increase in Dorfman pooling false-negative rates
    $\err{exact}$. Left: Colors represent $\err{exact}$, the relative
    percentage increase in the scheme false-negative rates relative to
    the single test false-negative rates equation \eqref{eq:erel}. Right:
    colors represent the difference between $\err{common}$ and
    $\err{exact}$. The infection prevalence $q$, is varied on the
    x-axis, while the test sensitivity is varied on the y-axis. Pool
    size $N$, was chosen according to $q$ as in
    \cite{DorfmanYuvalDor}. The left panel shows that $\err{exact}$ is
    largest for low prevalence values $q$. The difference between
    $\err{common}$ and $\err{exact}$ can be as large as 30\%, as seen
    in the right panel. }\label{fig1}
\end{figure}

@@

Furthermore, an especially important consideration in designing
pooling schemes is explicitly taking into account the intrinsic RT-PCR error rates
. As we have demonstrated above, analyzing the expected
error rates of pooling schemes under a realistic error
model is readily achievable. Designing pooling schemes that optimize
testing while accounting for error rates is, however, considerably more involved.
Nonetheless, various algorithms explicitly incorporate error
rates into their pooling optimization processes \cite{Kim, OptimalDorfmanPool}, including a
recent study by our group which employed the herein described error
model \cite{DOPE}.  In our proposed algorithm, we have used an
experimental design criterion to create an algorithm which iteratively
performs pooling, updating its information based on previous
results. In doing so, our pooling scheme shows potential in terms of
both reducing false-negative and false-positive rates, as well as increasing efficiency,
although it has not yet been implemented in real-world settings.
